---
title: "STAT 432 project"
output: html_document
---
DATA PREPARATION :
Reference : https://www.kaggle.com/datasets/samdemharter/brca-multiomics-tcga/versions/3?select=data.csv 
# IMPORT DATA
```{r}
brca = read.csv("brca_data_w_subtypes.csv")
# PR.Status
brca_pr = subset(brca, PR.Status == "Positive" | PR.Status == "Negative", select= -c(vital.status, ER.Status, HER2.Final.Status, histological.type))
```

```{r}
# Changing "Positive" to 1, "Negative" to 0
brca_pr[brca_pr["PR.Status"] == "Positive","PR.Status"] = 1
brca_pr[brca_pr["PR.Status"] == "Negative","PR.Status"] = 0
```

```{r}
# ER.Status
brca_er = subset(brca, ER.Status == "Positive" | ER.Status == "Negative", select= -c(vital.status, PR.Status, HER2.Final.Status, histological.type))
```

# Changing "Positive" to 1, "Negative" to 0
```{r}
brca_er[brca_er["ER.Status"] == "Positive","ER.Status"] = 1
brca_er[brca_er["ER.Status"] == "Negative","ER.Status"] = 0
```
# HER2.Final.Status
```{r}
brca_her2 = subset(brca, HER2.Final.Status == "Positive" | HER2.Final.Status == "Negative", select= -c(vital.status, ER.Status, PR.Status, histological.type))
```

# Changing "Positive" to 1, "Negative" to 0
```{r}
brca_her2[brca_her2["HER2.Final.Status"] == "Positive","HER2.Final.Status"] = 1
brca_her2[brca_her2["HER2.Final.Status"] == "Negative","HER2.Final.Status"] = 0
```


# histological.type
```{r}
brca_hist = subset(brca, histological.type == "infiltrating ductal carcinoma" | histological.type == "infiltrating lobular carcinoma", select= -c(vital.status, ER.Status, HER2.Final.Status, PR.Status))
```

# Changing "infiltrating ductal carcinoma" to 1, "infiltrating lobular carcinoma" to 0
```{r}
brca_hist[brca_hist["histological.type"] == "infiltrating ductal carcinoma","histological.type"] = 1
brca_hist[brca_hist["histological.type"] == "infiltrating lobular carcinoma","histological.type"] = 0
```




Missing value checking : 
```{r}
table(as.matrix(colSums(is.na(brca))))
```



Categorical Variable 
mu_:
```{r}
mu_data = brca[,1465:1713]
mu_names = colnames(mu_data)
count_0 = rep(NA, 249)
count_1 = rep(NA, 249)

freq_mu = data.frame(mu_names, count_0, count_1)
counts = table(mu_data[1])

for (i in 1:nrow(freq_mu)){
  counts = table(mu_data[i])
  freq_mu[i, "count_0"] = counts["0"]
  freq_mu[i, "count_1"] = counts["1"]
}
```


cn_:
```{r}
cn_data = brca[,605:1464]
cn_names = colnames(cn_data)
count_n2 = rep(NA, 860)
count_n1 = rep(NA, 860)
count_0 = rep(NA, 860)
count_p1 = rep(NA, 860)
count_p2 = rep(NA, 860)

freq_cn = data.frame(cn_names, count_n2, count_n1, count_0, count_p1, count_p2)

for (i in 1:nrow(freq_cn)){
  counts = table(cn_data[i])
  freq_cn[i, "count_n2"] = counts["-2"]
  freq_cn[i, "count_n1"] = counts["-1"]
  freq_cn[i, "count_0"] = counts["0"]
  freq_cn[i, "count_p1"] = counts["1"]
  freq_cn[i, "count_p2"] = counts["2"]
}

```


# replacing NAs with 0 for convenience
```{r}
freq_cn[is.na(freq_cn)] = 0
```


Continuous Variable
rs_:

```{r}
rs_data = brca[,1:604]
rs_names = colnames(rs_data)

rs_outliers = data.frame(matrix(NA, nrow = 604, ncol = 160))
rs_outliers["rs_names"] = rs_names
rs_outliers = rs_outliers[,c(161,1:160)]
```




## complicated version
```{r}
for (i in 1:nrow(rs_outliers)) {
 outliers = boxplot.stats(brca[,i])$out
 if (length(outliers) == 0){
   rs_outliers[i,2] = 0
 } else{
   for (j in 1:length(outliers)){
      rs_outliers[i,1+j] = outliers[j]
  }
 }
}

```


## count (simplified version)
```{r}
counts = rep(NA, 604)
rs_outliers_simplified = data.frame(rs_names, counts)

for (i in 1:nrow(rs_outliers_simplified)) {
 outliers = boxplot.stats(brca[,i])$out
 rs_outliers_simplified[i,2] = length(outliers)
}

```


pp_: number of outliers for each protein level
#pp
```{r}
pp_data = brca[,1714:1936]
pp_names = colnames(pp_data)

counts_pp = rep(NA, 223)
pp_outliers_simplified = data.frame(pp_names, counts_pp)


for (i in 1:nrow(pp_outliers_simplified)) {
 outliers = boxplot.stats(brca[,i+1713])$out
 pp_outliers_simplified[i,2] = length(outliers)
}
```


## Modeling for PR.status (svm method)
```{r}
brca_pr = subset(brca, PR.Status == "Positive" | PR.Status == "Negative", select= -c(vital.status, ER.Status, HER2.Final.Status, histological.type))

brca_pr[brca_pr["PR.Status"] == "Positive","PR.Status"] = 1
brca_pr[brca_pr["PR.Status"] == "Negative","PR.Status"] = 0

cost.grid = expand.grid(C = seq(0.01, 1, length = 10))
train_control = trainControl(method="cv", number=10)
svm2 <- train(as.factor(PR.Status)~., data =brca_pr, method = "svmLinear", trControl = train_control, tuneGrid = cost.grid)
testpred = predict(svm2,  newdata = brca_pr)
table(testpred, as.factor(brca_pr$PR.Status))

```


## Modeling for PR.status (lda method)
```{r}
library(MASS)
x= data.matrix(brca_pr[, -1937])
y= brca_pr$PR.Status
dig.lda=lda(x,y)
Ytest.pred = predict(dig.lda, x)
table(y,Ytest.pred$class)

```


## Modeling for Histological.Type (Logistic regression with Elastic Net penalty)
# histological.type
```{r}
library(glmnet)
library(ROCR)

## Logistic regression
set.seed(1)
# use cv.glmnet
alphas = seq(0.1,0.9,0.1)
cvm = matrix(NA, nrow = 9, ncol = 2)
for (i in 1:9){
  cvm[i,1] = alphas[i]
  cv_glm = cv.glmnet(x = data.matrix(brca_hist[,1:1936]),
                         y = brca_hist[,1937],
                         nfolds = 10,
                         type.measure = "auc",
                         family = "binomial",
                         alpha = alphas[i])
  cvm[i,2] = min(cv_glm$cvm)
}

#alpha value with the lowest cvm
cvm[which.min(cvm[,2])]

elst_glm = cv.glmnet(x = data.matrix(brca_hist[,1:1936]),
                         y = brca_hist[,1937],
                         nfolds = 10,
                         type.measure = "auc",
                         family = "binomial",
                         alpha = 0.8)

pred = predict(elst_glm, newx = as.matrix(brca_hist[,1:1936]), type = "response")

# confusion table
ctab = table(pred > 0.7, brca_hist$histological.type)
(ctab[2,1] + ctab[1,2])/sum(ctab)

# found out that cutoff of 0.7 gives the lowest classification error
# Unsure which classification error to use

# auc
roc = prediction(pred, brca_hist$histological.type)
pfm = performance(roc, "tpr", "fpr")
plot(pfm, colorize = TRUE)

 # Area under the curve
performance(roc, measure = "auc")@y.values
```


## Modeling for Histological.Type (KNN method)

```{r}
library(class)
#KNN tuning K
set.seed(1)
train_id = sample(430, replace = FALSE)
train_hist = brca_hist[train_id,]
test_hist = brca_hist[-train_id,]

# running For loop to find the k that gives highest AUC
k_values = seq(110,118,1)
auc_values = rep(NA, 9)
k_results = data.frame(k_values, auc_values)

for (i in 1:length(k_values)){
  knn_fit_prob = knn(train_hist[,-1937], test_hist[,-1937],train_hist[,1937], k = k_values[i], prob = TRUE)
  prob = attr(knn_fit_prob, "prob")
  roc = prediction(prob, test_hist$histological.type)
  k_results[i,2] = performance(roc, measure = "auc")@y.values[[1]]
}

## k value with the highest AUC
k_results[which.max(k_results$auc_values),]

## Refit knn model using the selected k value

knn_fit = knn(train_hist[,-1937], test_hist[,-1937],train_hist[,1937], k = 114, prob = TRUE)
prob = attr(knn_fit, "prob")
roc = prediction(prob, test_hist$histological.type)
performance(roc, measure = "auc")@y.values[[1]]
pfm2 = performance(roc, "tpr", "fpr")
plot(pfm2, colorize = TRUE)
```


## VARIABLE SELECTION
```{r}
set.seed(1)
cv_group = sample(1:3, 705, replace = TRUE)
brca["ID"] = cv_group
# subset data

# PR.Status
brca_pr = subset(brca, PR.Status == "Positive" | PR.Status == "Negative", select= -c(vital.status, ER.Status, HER2.Final.Status, histological.type))
# Changing "Positive" to 1, "Negative" to 0
brca_pr[brca_pr["PR.Status"] == "Positive","PR.Status"] = 1
brca_pr[brca_pr["PR.Status"] == "Negative","PR.Status"] = 0

# ER.Status
brca_er = subset(brca, ER.Status == "Positive" | ER.Status == "Negative", select= -c(vital.status, PR.Status, HER2.Final.Status, histological.type))
# Changing "Positive" to 1, "Negative" to 0
brca_er[brca_er["ER.Status"] == "Positive","ER.Status"] = 1
brca_er[brca_er["ER.Status"] == "Negative","ER.Status"] = 0

# HER2.Final.Status
brca_her2 = subset(brca, HER2.Final.Status == "Positive" | HER2.Final.Status == "Negative", select= -c(vital.status, ER.Status, PR.Status, histological.type))
# Changing "Positive" to 1, "Negative" to 0
brca_her2[brca_her2["HER2.Final.Status"] == "Positive","HER2.Final.Status"] = 1
brca_her2[brca_her2["HER2.Final.Status"] == "Negative","HER2.Final.Status"] = 0


# histological.type
brca_hist = subset(brca, histological.type == "infiltrating ductal carcinoma" | histological.type == "infiltrating lobular carcinoma", select= -c(vital.status, ER.Status, HER2.Final.Status, PR.Status))
# Changing "infiltrating ductal carcinoma" to 1, "infiltrating lobular carcinoma" to 0
brca_hist[brca_hist["histological.type"] == "infiltrating ductal carcinoma","histological.type"] = 1
brca_hist[brca_hist["histological.type"] == "infiltrating lobular carcinoma","histological.type"] = 0
```

## 50 VARIABLES SELECTED 
```{r}
var_names =  c("rs_ESR1", "rs_FSIP1","rs_TFF1","rs_AGR3","rs_AGR2","rs_BCAS1","rs_FOXA1","rs_NAT1","cn_NAT1","mu_FOXA1","cn_FOXA1","mu_GATA3","cn_GATA3","mu_TBX3","mu_PTEN","mu_TP53","pp_LKB1","mu_CDH1","cn_CDH1","mu_ATM","pp_ATM","mu_MAP3K1","pp_BRCA2","rs_PGR","cn_PGR", "pp_XBP1","mu_ERBB3","rs_ERBB4","pp_EGFR","cn_MELK","rs_MMP11","cn_MMP11","rs_TMEM45B","cn_TMEM45B","rs_KRT17", "cn_KRT17","rs_KRT5","rs_SFRP1","cn_SFRP1","rs_KRT14","cn_KRT14","rs_APOD","cn_FGFR4","rs_MIA","cn_MIA","rs_DHRS2","cn_FOXC1","rs_HMGCS2","cn_CDH3","cn_MAPT")
length(var_names)
```


## PREDICTING PR.STATUS
```{r}
set.seed(1)
# PR.Status

brca_pr_50 = brca_pr[,c(var_names,"PR.Status","ID")]

# test AUC

alphas = seq(0.1,0.9,0.1)
avg_auc = rep(NA, 3)

cvm = matrix(NA, nrow = 9, ncol = 2)

for (i in 1:3){
  fold_data_train = subset(brca_pr_50, ID == i, select= -c(ID))
  fold_data_test = subset(brca_pr_50, ID != i, select= -c(ID))
  for (j in 1:9){
    cvm[j,1] = alphas[j]
    cv_glm = cv.glmnet(x = data.matrix(fold_data_train[,1:50]),
                         y = fold_data_train[,51],
                         nfolds = 10,
                         type.measure = "auc",
                         family = "binomial",
                         alpha = alphas[j])
    cvm[j,2] = mean(cv_glm$cvm)
  }
  elst_glm = cv.glmnet(x = data.matrix(fold_data_train[,1:50]),
                         y = fold_data_train[,51],
                         nfolds = 10,
                         type.measure = "auc",
                         family = "binomial",
                         alpha = cvm[which.min(cvm[,2])])
pred = predict(elst_glm, newx = as.matrix(fold_data_test[,1:50]), type = "response")
roc = prediction(pred, fold_data_test$PR.Status)

 # Area under the curve
avg_auc[i] = performance(roc, measure = "auc")@y.values
}

(avg_auc[[1]] + avg_auc[[2]] + avg_auc[[3]])/3
```


## PREDICTING ER.STATUS
```{r}
# ER.Status
set.seed(1)
brca_er_50 = brca_er[,c(var_names,"ER.Status","ID")]

# test AUC

alphas = seq(0.1,0.9,0.1)
avg_auc = rep(NA, 3)

cvm = matrix(NA, nrow = 9, ncol = 2)

for (i in 1:3){
  fold_data_train = subset(brca_er_50, ID == i, select= -c(ID))
  fold_data_test = subset(brca_er_50, ID != i, select= -c(ID))
  for (j in 1:9){
    cvm[j,1] = alphas[j]
    cv_glm = cv.glmnet(x = data.matrix(fold_data_train[,1:50]),
                         y = fold_data_train[,51],
                         nfolds = 10,
                         type.measure = "auc",
                         family = "binomial",
                         alpha = alphas[j])
    cvm[j,2] = mean(cv_glm$cvm)
  }
  elst_glm = cv.glmnet(x = data.matrix(fold_data_train[,1:50]),
                         y = fold_data_train[,51],
                         nfolds = 10,
                         type.measure = "auc",
                         family = "binomial",
                         alpha = cvm[which.min(cvm[,2])])
pred = predict(elst_glm, newx = as.matrix(fold_data_test[,1:50]), type = "response")
roc = prediction(pred, fold_data_test$ER.Status)

 # Area under the curve
avg_auc[i] = performance(roc, measure = "auc")@y.values
}

(avg_auc[[1]] + avg_auc[[2]] + avg_auc[[3]])/3
```


## PREDICTING HER2.FINAL.STATUS
```{r}
# HER2
set.seed(1)
brca_her2_50 = brca_her2[,c(var_names,"HER2.Final.Status","ID")]

# test AUC

alphas = seq(0.1,0.9,0.1)
avg_auc = rep(NA, 3)

cvm = matrix(NA, nrow = 9, ncol = 2)

for (i in 1:3){
  fold_data_train = subset(brca_her2_50, ID == i, select= -c(ID))
  fold_data_test = subset(brca_her2_50, ID != i, select= -c(ID))
  for (j in 1:9){
    cvm[j,1] = alphas[j]
    cv_glm = cv.glmnet(x = data.matrix(fold_data_train[,1:50]),
                         y = fold_data_train[,51],
                         nfolds = 10,
                         type.measure = "auc",
                         family = "binomial",
                         alpha = alphas[j])
    cvm[j,2] = mean(cv_glm$cvm)
  }
  elst_glm = cv.glmnet(x = data.matrix(fold_data_train[,1:50]),
                         y = fold_data_train[,51],
                         nfolds = 10,
                         type.measure = "auc",
                         family = "binomial",
                         alpha = cvm[which.min(cvm[,2])])
pred = predict(elst_glm, newx = as.matrix(fold_data_test[,1:50]), type = "response")
roc = prediction(pred, fold_data_test$HER2.Final.Status)

 # Area under the curve
avg_auc[i] = performance(roc, measure = "auc")@y.values
}

(avg_auc[[1]] + avg_auc[[2]] + avg_auc[[3]])/3
```

## PREDICTING HISTOLOGICAL.TYPE
```{r}
# histological.type
set.seed(1)
brca_hist_50 = brca_hist[,c(var_names,"histological.type","ID")]

# test AUC

alphas = seq(0.1,0.9,0.1)
avg_auc = rep(NA, 3)

cvm = matrix(NA, nrow = 9, ncol = 2)

for (i in 1:3){
  fold_data_train = subset(brca_hist_50, ID == i, select= -c(ID))
  fold_data_test = subset(brca_hist_50, ID != i, select= -c(ID))
  for (j in 1:9){
    cvm[j,1] = alphas[j]
    cv_glm = cv.glmnet(x = data.matrix(fold_data_train[,1:50]),
                         y = fold_data_train[,51],
                         nfolds = 10,
                         type.measure = "auc",
                         family = "binomial",
                         alpha = alphas[j])
    cvm[j,2] = mean(cv_glm$cvm)
  }
  elst_glm = cv.glmnet(x = data.matrix(fold_data_train[,1:50]),
                         y = fold_data_train[,51],
                         nfolds = 10,
                         type.measure = "auc",
                         family = "binomial",
                         alpha = cvm[which.min(cvm[,2])])
pred = predict(elst_glm, newx = as.matrix(fold_data_test[,1:50]), type = "response")
roc = prediction(pred, fold_data_test$histological.type)

 # Area under the curve
avg_auc[i] = performance(roc, measure = "auc")@y.values
}

(avg_auc[[1]] + avg_auc[[2]] + avg_auc[[3]])/3
```

RESULTS:
PR.Status: 0.9024917
ER.Status: 0.9463468
HER2.Final.Status: 0.6845975
histological.type: 0.8774072

Average: 0.85269212













